Hurricane Intensification and SST
=================================

### James B. Elsner and Erik Fraza

Hourly best track data as a data frame
```{r}
library("dplyr")
load("best.use.2014.RData")
begin = 1986; end = 2013
Tracks.df = best.use %>%
  mutate(Int = WmaxS * .5144, 
         DIntDt = DWmaxDt * .5144) %>%
  filter(Yr >= begin, Yr <= end, Int >= 33, M == FALSE)
```

Distribution of intensity change
```{r}
library("ggplot2")
library("scales")
ggplot(Tracks.df, aes(DIntDt)) +
  geom_histogram(binwidth = .5, color = "white", fill = muted("blue")) +
  xlab(expression(paste("Intensity Change [m ", s^-1, " per h]"))) +
  ylab("Number of Hours") +
  geom_histogram(aes(DIntDt), data = Tracks.df[Tracks.df$DIntDt >= 0, ], 
                 binwidth = .5, color = "white", fill = "#6A51A3") +
  scale_x_continuous(limits = c(-4, 4)) 
length(Tracks.df$DIntDt)
length(unique(Tracks.df$Sid))
mean(Tracks.df$DIntDt)
fivenum(Tracks.df$DIntDt)
library("moments")
kurtosis(Tracks.df$DIntDt)
```

```{r}
Tracks.df = Tracks.df %>%
  filter(DIntDt > 0)
mean(Tracks.df$DIntDt)
median(Tracks.df$DIntDt)
median(Tracks.df$DIntDt) * 1.9438 * 24 #convert to kt/day
fivenum(Tracks.df$DIntDt)
ggplot(Tracks.df, aes(DIntDt)) +
  geom_histogram(binwidth = .25, color = "white", fill = muted("red"))
```

Track points as a spatial points data frame
```{r}
library("rgdal")
ll = "+proj=longlat +ellps=WGS84"
#lcc = "+proj=lcc +lat_1=60 +lat_2=30 +lon_0=-60 +units=km"
Tracks.sdf = Tracks.df
coordinates(Tracks.sdf) = c("lon", "lat")
proj4string(Tracks.sdf) = CRS(ll)
```

Create grid
```{r}
library("raster")
r = raster(ncol = 20, nrow = 10, 
           xmn = -100, xmx = -20, 
           ymn = 10, ymx = 50)
Example.sdf = subset(Tracks.sdf, 
                     Sid == 1517 | 
                     Sid == 1605 |
                     Sid == 1657)
Example.grid = rasterize(Example.sdf, r, 
                         field = 'DIntDt',
                         fun = mean)
```

Plot grids
```{r}
library("RColorBrewer")
library("rasterVis")
range(values(Example.grid), na.rm = TRUE)
rng = seq(0, 1.4, .2)
cr = brewer.pal(8, "Purples")
cr = cr[-1]
vals = levelplot(Example.grid, margin = FALSE, 
          sub = expression(paste("     Average Intensification [m ", s^-1, " per h]")), 
          xlab = NULL, ylab = NULL, 
          col.regions = cr, at = rng, 
          colorkey = list(space = 'bottom'),
          border = "white", border.lwd = 2,
          par.settings = list(fontsize = list(text = 15)))
```

Country borders
```{r}
library(mapproj)
library(maptools)
outlines = as.data.frame(map("world", xlim = c(-100, -20), 
                             ylim = c(10, 50), 
                             plot = FALSE)[c("x", "y")],
                             color = "gray")
map = geom_path(aes(x, y), inherit.aes = FALSE, data = outlines, 
                alpha = .8, show_guide = FALSE, color = "gray")
ext = as.vector(extent(r))
boundaries = map("world", fill = TRUE, xlim = ext[1:2], 
                 ylim = ext[3:4], plot = FALSE)
IDs = sapply(strsplit(boundaries$names, ":"), function(x) x[1])
bPols = map2SpatialPolygons(boundaries, IDs = IDs,
                              proj4string = CRS(projection(r)))
```

Tracks
```{r}
require(gridExtra)
Int2a.df = filter(Tracks.df, Sid == 1517)
Int2b.df = filter(Tracks.df, Sid == 1605)
Int2c.df = filter(Tracks.df, Sid == 1657)
holdera = data.frame(Int2a.df$lon, Int2a.df$lat)
holderb = data.frame(Int2b.df$lon, Int2b.df$lat)
holderc = data.frame(Int2c.df$lon, Int2c.df$lat)
linesa = SpatialLines(list(Lines(list(Line(holdera)),
                                 ID = 'Int2a.df$Sid')))
linesb = SpatialLines(list(Lines(list(Line(holderb)),
                                 ID = 'Int2b.df$Sid')))
linesc = SpatialLines(list(Lines(list(Line(holderc)),
                                 ID = 'Int2c.df$Sid')))
vals + layer(sp.polygons(bPols, col = gray(.8))) +
  layer(sp.lines(linesa, lwd = 2, col = gray(.4))) + 
  layer(sp.lines(linesb, lwd = 2, col = gray(.4))) + 
  layer(sp.lines(linesc, lwd = 2, col = gray(.4))) +
  layer(panel.text(-64, 27.5, 'Floyd (1999)', col = gray(.4))) +
  layer(panel.text(-48, 15.5, 'Dean (2007)', col = gray(.4))) + 
  layer(panel.text(-79.75, 12, 'Ivan (2004)', col = gray(.4)))
```

Mean intensification grid
```{r}
Intfy.grid = rasterize(Tracks.sdf, r, 
                     field = 'DIntDt',
                     fun = mean)
sum(!is.na(values(Intfy.grid)))
range(values(Intfy.grid), na.rm = TRUE)
rng = seq(0, 1.2, .2)
cr = brewer.pal(8, "Purples")
cr = cr[-1]
cr = cr[-length(cr)]
vals = levelplot(Intfy.grid, margin = FALSE, 
          sub = expression(paste("             Mean Intensification [m ", s^-1, " per h]")), 
          xlab = NULL, ylab = NULL, 
          col.regions = cr, at = rng, 
          colorkey = list(space = 'bottom'),
          border = "white", border.lwd = 2,
          par.settings = list(fontsize = list(text = 15)))
pA = vals + layer(sp.polygons(bPols, col = gray(.8)))
pA = update(pA, main = textGrob("a", x = unit(.05, "npc"), 
                               gp = gpar(fontsize=16)))

Freq.grid = rasterize(Tracks.sdf, r, 
                     field = 'DIntDt',
                     fun = 'count')
SD.grid = rasterize(Tracks.sdf, r,
                    field = 'DIntDt',
                    fun = sd)
SE.grid = SD.grid/sqrt(Freq.grid)
CV.grid = SD.grid/Intfy.grid

range(values(CV.grid), na.rm = TRUE)
rng = seq(0, 2, .5)
cr = brewer.pal(6, "Greens")
cr = cr[-1]
vals = levelplot(CV.grid, margin = FALSE, 
          sub = expression(paste("            Coefficient of Variation")), 
          xlab = NULL, ylab = NULL, 
          col.regions = cr, at = rng, 
          colorkey = list(space = 'bottom'),
          border = "white", border.lwd = 2,
          par.settings = list(fontsize = list(text = 15)))
pB = vals + layer(sp.polygons(bPols, col = gray(.8)))
pB = update(pB, main = textGrob("b", x = unit(.05, "npc"), 
                               gp = gpar(fontsize = 16)))

grid.arrange(pA, pB, ncol = 1)
```

SST data
```{r}
load("ncSST20002013.RData")  
load("ncSST19901999.RData")
load("ncSST19801989.RData")

SST = cbind(ncdataframe19801989[, -(3:74)], 
            ncdataframe19901999[, 3:length(ncdataframe19901999)], 
            ncSST20002013[, 3:length(ncSST20002013)])
Month = rep(month.abb, length(1986:2013))
Index = Month == "Jun" | Month == "Jul" | Month == "Aug" | Month == "Sep" | Month == "Oct"
SST2 = SST[, which(Index) + 2]
sst = rowMeans(SST2)
sst.df = data.frame(lon = SST$lon, lat = SST$lat, sst = sst)
sst.sdf = sst.df
coordinates(sst.sdf) = c("lon", "lat")
proj4string(sst.sdf) = CRS(ll)
```

```{r}
SST.grid = rasterize(sst.sdf, r,
                     field = 'sst', 
                     fun = mean)
SST.grid = mask(SST.grid, Intfy.grid)
sum(!is.na(values(SST.grid)))

SST.grid2 = SST.grid
SST.grid2[SST.grid2 < 20] = 20

range(values(SST.grid), na.rm = TRUE)
rng = seq(20, 30, 2)
cr = brewer.pal(6, "Reds")
cr = cr[-1]
vals = levelplot(SST.grid2, margin = FALSE, 
          sub = expression(paste("          Averge SST [", degree, "C]")), 
          xlab = NULL, ylab = NULL, 
          col.regions = cr, at = rng, 
          colorkey = list(space = 'bottom'),
          border = "white", border.lwd = 2,
          par.settings = list(fontsize = list(text = 15)))
vals + layer(sp.polygons(bPols, col = gray(.8)))
```

Additional grids
```{r}
Speed.grid = rasterize(Tracks.sdf, r, 
                     field = 'maguv',
                     fun = mean)
Wmax.grid = rasterize(Tracks.sdf, r,
                      field = 'Wmax',
                      fun = mean)

cellStats(Speed.grid, mean)
m = Speed.grid
m[m < 22] = NA
mr = mask(Speed.grid, m)
cellStats(mr, mean) * .5144
m = Speed.grid
m[m >= 22] = NA
mr = mask(Speed.grid, m)
cellStats(mr, mean) * .5144
```

Crop extent and create data frame
```{r}
ext = extent(Intfy.grid, 4, 10, 2, 17)
Intfy.gridC = crop(Intfy.grid, ext)
SST.gridC = crop(SST.grid, ext)
Speed.gridC = crop(Speed.grid, ext)
Wmax.gridC = crop(Wmax.grid, ext)
Freq.gridC = crop(Freq.grid, ext)
SE.gridC = crop(SE.grid, ext)

int = values(Intfy.gridC)
sst = values(SST.gridC)
speed = values(Speed.gridC)
count = values(Freq.gridC)
Wmax = values(Wmax.gridC)
seInt = values(SE.gridC)
Loc = coordinates(Intfy.gridC)

df = data.frame(x = Loc[, 1], y = Loc[, 2], int, seInt, sst, speed, count, Wmax)
summary(df)
summary(df$int) * 1.9438 * 24

Mask = SST.gridC
Mask[Mask > 0] = FALSE

cr = gray(.8)
vals = levelplot(Mask, margin = FALSE,
          xlab = NULL, ylab = NULL, 
          col.regions = cr, 
          colorkey = FALSE,
          border = "white", border.lwd = 2)
vals + layer(sp.polygons(bPols, col = gray(.5)))
```

Histogram of mean intensification
```{r}
ggplot(df, aes(int)) +
  geom_histogram(binwidth = .1, color = "white", fill = "#6A51A3") +
  xlab(expression(paste("Average Intensification [m ", s^-1, " per h]"))) +
  ylab("Number of Grid Cells")
```

Intensification vs SST
```{r}
cor.test(df$sst, df$int)
cor.test(df$sst, df$Wmax)
cor.test(df$int, df$count)
cor.test(df$y, df$sst)

model = glm(int ~ sst + Wmax, family = Gamma(link = log), data = df)
summary(model)
confint(model)

summary(df$int, na.rm = TRUE) * 1.9438 * 24
summary(df$sst, na.rm = TRUE)
```

Scatter plot and model at mean Wmax
```{r}
pred = predict(model, newdata = data.frame(int = df$int, sst = df$sst, Wmax = mean(df$Wmax, na.rm = TRUE)), 
               type = "response", se.fit = TRUE)
df$pred = pred$fit
df$predSE = pred$se.fit
df$residuals = df$int - df$pred

range(df$residuals, na.rm = TRUE)
median(df$residuals, na.rm = TRUE)
ggplot(df, aes(x = sst, y = int)) + 
  geom_line(aes(x = sst, y = pred), size = 1, color = "purple") +
  geom_ribbon(aes(x = sst, ymax = pred + 2 * predSE, ymin = pred - 2 * predSE), 
              color = "gray", alpha = .25) +
  geom_point() +
  geom_errorbar(aes(ymin = int - seInt, ymax = int + seInt), width = .05) +
  xlab(expression(paste("Average SST [", degree, "C]")))+
  ylab(expression(paste("Mean Intensification [m " ~ s^{-1} ~ " per h]")))
```

Residuals
```{r}
pA = ggplot(df, aes(residuals)) +
  geom_histogram(binwidth = .1, color = "white", fill = gray(.75)) +
  xlab(expression(paste("Residual [m ", s^-1, " per h]"))) +
  ylab("Number of Grid Cells")

qqplot.data <- function (vec) # argument: vector of numbers
{
  # following four lines from base R's qqline()
  y <- quantile(vec[!is.na(vec)], c(0.25, 0.75))
  x <- qnorm(c(0.25, 0.75))
  slope <- diff(y)/diff(x)
  int <- y[1L] - slope * x[1L]

  d <- data.frame(resids = vec)

  ggplot(d, aes(sample = resids)) +
    geom_abline(slope = slope, intercept = int, color = 'purple') +
    stat_qq(size = 3, color = gray(.75)) + 
    xlab("Theoretical Quantiles") +
    ylab(expression(paste("Residual Quantiles [m " ~ s^{-1} ~ " per h]")))
}
pB = qqplot.data(df$residuals)

source("multiplot.txt")
mat = matrix(c(1, 2), nrow = 1, byrow = TRUE)
pA = pA + ggtitle("a") + theme(plot.title = element_text(hjust = 0))
pB = pB + ggtitle("b") + theme(plot.title = element_text(hjust = 0))              
multiplot(pA, pB, layout = mat)
```

Map of residuals
```{r}
Resid.grid = crop(r, ext)
Resid.grid[] = df$residuals
range(values(Resid.grid), na.rm = TRUE)
median(values(Resid.grid), na.rm = TRUE)
rng = seq(-.8, .8, .2)
cr = brewer.pal(8, "BrBG")
vals = levelplot(Resid.grid, margin = FALSE, 
          sub = expression(paste("        Residuals [m " ~ s^{-1} ~ "" ~ h^{-1} ~ "per ",degree,"C]")), 
          xlab = NULL, ylab = NULL, 
          col.regions = cr, at = rng, 
          colorkey = list(space = 'bottom'),
          border = "white", border.lwd = 2,
          par.settings = list(fontsize = list(text = 15)))
vals + layer(sp.polygons(bPols, col = gray(.8)))
```

Spatial autocorrelation

Convert to spatial polygons data frame
```{r}
X = stack(Intfy.gridC, SST.gridC, Wmax.gridC)
X.spdf = rasterToPolygons(X)
names(X.spdf) = c("int", "sst", "Wmax")
X.spdf$ID = 1:dim(X.spdf@data)[1]
```

Neighborhood definition
```{r}
library("spdep")
nb = poly2nb(X.spdf, queen = TRUE)
wts = nb2listw(nb)
```

Test for spatial autocorrelation in the residuals
```{r}
model = glm(int ~ sst + Wmax, family = Gamma(link = log), data = X.spdf@data)
summary(model)
lm.morantest(model, wts)
moran.test(X.spdf$int, wts)
```

Local spatial autocorrelation
```{r}
#LM.grid = MoranLocal(Resid.grid, w = matrix(rep(0:1, length = 9), 3, 3))
LM.grid = MoranLocal(Resid.grid)
range(values(LM.grid), na.rm = TRUE)
rng = seq(-1, 7, 1)
cr1 = brewer.pal(9, "Oranges")
cr2 = brewer.pal(9, "YlGn")
cr = c(cr2[4], cr1[3:9])
vals = levelplot(LM.grid, margin = FALSE, 
          sub = expression(paste("        Residual Spatial Correlation")), 
          xlab = NULL, ylab = NULL, 
          col.regions = cr, at = rng, 
          colorkey = list(space = 'bottom'),
          border = "white", border.lwd = 2,
          par.settings = list(fontsize = list(text = 15)))
vals + layer(sp.polygons(bPols, col = gray(.8)))
```

```{r}
TT.df = Tracks.df %>%
  filter(lon <= -64, lon >= -76, lat >= 10, lat <= 14)
unique(TT.df$name)
unique(TT.df$Yr)
```

Residuals correlated with other variables
```{r}
X = stack(Resid.grid, Speed.gridC)
nlayers(X)
layerStats(X, 'pearson', na.rm = TRUE)
```

Spatial Regression

Spatial error or spatial lag model.
```{r}
modelLM = lm(int ~ sst + Wmax, data = X.spdf@data)
summary(modelLM)
lm.LMtests(modelLM, wts, test = c("LMerr","LMlag"))
lm.LMtests(modelLM, wts, test = c("RLMerr","RLMlag"))
```

INLA model
```{r}
library("INLA")
nb2INLA("graph.inla", nb)
graph.inla = inla.read.graph("graph.inla")
formula = int ~ sst + Wmax + f(ID, model = "besag", graph = graph.inla)
model.sr = inla(formula = formula, family = "gamma",
                 data = X.spdf@data, 
                 control.compute = list(config = TRUE, mlik = TRUE, cpo = TRUE, dic = TRUE, po = TRUE),
                 control.predictor = list(compute = TRUE),
                 control.results = list(return.marginals.random = TRUE),
                 control.family = list(hyper = list(theta = list(prior = "loggamma", param = c(1, 1))))
                 )
summary(model.sr)

```

```{r}
X.spdf$random = model.sr$summary.random$ID$mean
spplot(X.spdf, "random")

X.spdf$fitted = model.sr$summary.fitted.values$mean
spplot(X.spdf, "fitted")
```


