---
title: "Hurricane Intensification"
author: "James B. Elsner"
date: October 13, 2018
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Import the data. The data are interpolated to one hour intervals. See Dropbox/Hurricanes/BestTrack.
```{r}
df <- read.csv(file = "best.interp.2016.csv", header = TRUE)
```

The variable `WmaxS` has units of knots consistent with the best-track wind speeds and the intensification rate (`DWmaxDt`) has units of knots per hour. Multiply by .5144 to get m/s and m/s per hour. Only consider hurricanes over water (`M == FALSE`).
```{r}
library(dplyr)
begin <- 1980
end <- 2016
Tracks.df <- df %>%
  mutate(Int = WmaxS * .5144, 
         DIntDt = DWmaxDt * .5144) %>%
  filter(Yr >= begin, Yr <= end, Int >= 33, M == FALSE)
```

Distribution of hourly intensification rates. Rapid intensification 30 kt/24 hr reduced amounts to .643 m/s per hr
```{r}
library(ggplot2)
library(scales)
ggplot(Tracks.df, aes(x = DIntDt)) +
  geom_freqpoly(bins = 50, size = 2) +
  geom_vline(xintercept = .643, color = "blue", size = 1.5) +
  geom_vline(xintercept = 0, color = "gray", size = .5) +
  xlab(expression(paste("Intensity Change [m ", s^-1, " per h]"))) +
  ylab("Number of Hours") +
  theme_minimal() +
  labs(title = "Blue line: Avg hourly intensification needed for `rapid intensification`",
       subtitle = "Steady state is most common") +
  annotate("label", x = -2.8, y = 200, label = "Data: NHC (1986-2016)") +
  annotate("label", x = 2.5, y = 2300, label = "Interpolation method\n (Elsner & Jagger 2013)")
```

Percentage of all intensifying hurricane hours with rates exceeding `rapid intensification`.
```{r}
Tracks.df %>%
  filter(DIntDt > 0) %>%
  group_by(SYear) %>%
  summarize(nH = n(),
            perH = sum(DIntDt > .643)/nH) %>%
ggplot(., aes(x = SYear, y = perH)) +
  geom_point() +
  geom_smooth(method = lm)
```

Other information.
```{r}
length(Tracks.df$DIntDt)
length(unique(Tracks.df$Sid))
mean(Tracks.df$DIntDt)
fivenum(Tracks.df$DIntDt)
library(moments)
kurtosis(Tracks.df$DIntDt)
```
